---
title: "Introduction to bdrc"
author: "Axel Örn Jansson, Sölvi Rögnvaldsson and Birgir Hrafnkelsson"
date:
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8, 
  fig.height=5, 
  fig.path="figs-introduction/"
)
```
A discharge rating curve is a model that describes the relationship between water stage and discharge in a river. The rating curve is estimated from paired observations of stage and discharge. The rating curve is used to predict discharge given stage. This is the main practical usage of rating curves as stage is substantially easier to directly observe than discharge. Four different dishcarge rating curve models are implemented in this R package using a Bayesian hierarchical model:

* bplm0() - Power-law model with constant variance (hence the 0). This is a Bayesian hierarchical implementation of the most commonly used discharge rating curve model

* bplm() - Power-law model with variance that may vary with stage

* bgplm0() - Generalized power-law model with constant variance (hence the 0)

* bgplm() - Generalized power-law model with variance that may vary with stage

For further details about the different models, see x. The models differ in their complexity and the bgplm is the most flexible and complex model. We will focus on the use of bgplm throughout this introduction vignette and explore the different ways to fit the gplm and visualize its output. However the API of the functions for the other three models are completely identical so this vignette also helps users to run those models. 
```{r,message=F}
#load the packages needed for this vignette
library(bdrc)
library(ggplot2)
library(dplyr)
```
We will use a dataset from a river called x in Sweden that comes with the package:
```{r, data}
data(V316_river)
V316_river
```
## Fit a discharge rating curve
There are two mandatory input arguments, formula and data. The formula is of the form y~x where y is discharge in $m^3/s$ and x is stage in $m$ (it is very important that the data is in the correct units)


```{r}
bgplm.fit <- bgplm(Q~W,data=V316_river)
```


```{r, fig.width=6, fig.height=5}
plot(bgplm.fit)
```

## Assess model fit and convergence
```{r, fig.width=6, fig.height=5}
plot(bgplm.fit,type='residuals')
```

```{r, fig.width=6, fig.height=5}
plot(bgplm.fit,type='trace','hyperparameters',transformed=T)
```
## Comparing different models
